---
title: "Hg Isotopic Dilution Approaches for R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hg Isotopic Dilution Approaches for R}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

The HgIDAR package can be use to calculate Hg concentrations and Hg methylation and demathylation ratios using isotopic dilution approaches (Monperrus et al., 2005; Rodriguez-Gonzalez et al., 2013). 

## Installation

This package is available in github repository and installation can be done with the following command:

- First, you need to install the devtools package. You can do this from CRAN. Invoke R and then type

```{r, eval=FALSE}
install.packages("devtools")
```
- Load the devtools package.
```{r, eval=FALSE}
library(devtools)
```
- In most cases, you just use install_github("author/package"). For example, with my R/broman package, which exists at github.com/micro-hg-cpg/HgIDAR", you’d type

```{r, eval=FALSE}
install_github("micro-hg-cpg/HgIDAR", dependencies = TRUE, build_vignettes = TRUE)
```

```{r setup}
library(HgIDAR)
```

## Example data
Three examples are provided:
- Example 1 correspondes to water samples in which isotope 199 enriched inorganic mercury (IHg) and isotope 201 enriched monomethylmercury (MMHg) were used of incubation and isotope 198 enriched IHg and 202 MMHg were used for quantification. 
- Example 2 corresponds to sediment samples in which isotope 204 enriched inorganic mercury (IHg) and isotope 200 enriched monomethylmercury (MMHg) were used of incubation and isotope 199 enriched IHg and 201 MMHg were used for quantification.
- Example 3 corresponds also to sediment samples in which isotope 199 enriched inorganic mercury (IHg) and isotope 201 enriched monomethylmercury (MMHg) were used of incubation and isotope 198 enriched IHg and 202 MMHg were used for quantification. 

To explore the functions available in this package, you can use any of the  mock-up scenarios provided in the examples. We have choose to use example 1 in this explanation. 
This mock up experiment is a sea water that has been double spiked, with isotope 199 enriched inorganic mercury (IHg) and isotope 201 enriched monomethylmercury (MMHg) and then incubated. These incubation spikes will allow us to measure the transformations; by measuring the concentration of 199 MMHg, we can calculate how much MMHg has form during the incubation and thus the potential methylation rate constant. Likewise by measuring the concentration of 201 IHg, we can calculate how much MMHg has undergone demethylation and thus calculate the potencial demethylation rate constant.

In order to measure the concentrations of each isotope, we need to add an quantification spike (isotope enriched 198 IHg and 202 MMHg) to act as an internal reference.

We need several files in order to make the calculations:

-   The integrated peak areas of Hg isotopes. The results from the ICP-MS-GC need to be processed before hand and only the peaks areas will be used for this calculations.

```{r, example_peaks}
head(peaks_1)

```

-   The masses of enriched stable isotopes used for quantification and the mass of water sample. In the case of sediments or other solids, mass of sediment sample and acid used during the digestions and the mass of extract used for derivatization.

```{r, example_masses}
head(masses_1)
```

-   The  working abundance matrix. Relative abundance of Hg isotopes of the isotopic modified spikes used in the incubation and quantification plus at least one isotope with the natural abundance. The abundance matrix needs to have 5 columns. 

```{r, example_abundance}
abundances_1
```

We provided the necessary files to run this packages and we recommend to use them as templates for your own samples (this will facilitate the running of the package)


## Checking the variability of the integrations

The variability  of your peak integrations should be checked before starting the IDA calculations.
We have implemented two steps in the quality check. 
- Initially, the mean, sd and rsd of the peak's area are calculated. We can directly see the variation among the injections. The package provides two different commands to obtain this information, rsd_peaks() will provide a table with all information and plot_rsd_peaks() will provide plots with the rsd (y-axis) and sample number (x-axis).  

```{r, variability peak check}
step1_quality_verification_table <- rsd_peaks(peaks_1, form = 'all', cutoff = 2)
step1_quality_verification_plot <- plot_rsd_peaks(peaks_1, form = 'all', cutoff = 2)

```

- Secondly and crucially, we calculate the ratio between the isotopes of interest: the quantification and incubation isotopic modified spikes as well as the isotope selected for natural abundance. Once the ratios are calculated for each injection (analytic replicates), this package can calculate the relative standard deviation (rsd). The package provides two different commands to obtain this information, rsd_ratios() will provide a table with all ratios and their rsd and plot_rsd_ratios() will provide plots with the rsd incub/quant (y-axis) and sample number (x-axis).  


```{r, variability ratios check}
step2_quality_verification_table <- rsd_ratios(peaks_1, form = 'MMHg', sample = 24, incub_IHg = 199, incub_MMHg = 201, quant_IHg = 198, quant_MMHg = 202, nat = 200)

step2_quality_verification_plot <- plot_rsd_ratios(peaks_1, form = 'MMHg', sample = 24, incub_IHg = 199, incub_MMHg = 201, quant_IHg = 198, quant_MMHg = 202, nat = 200)

```

When the relative standard deviation among injections is too high, you need to go back to your integrations and recheck them before moving on. The default cutoff for the rsd is set to 10%, however this can be modified. 

##Calculate the concentration 

This package offers the possibility of calculating the concentrations of both IHg and MMHg either by injection or by sample. 

```{r, concentrations}
#Either calculating the individual values of each injection
IHg_concentrations_byinjection <- IHg(peaks_1, masses_1, abundances_1, 24, 3, type = 'liquid', incub_IHg = 199, incub_MMHg = 201, quant_IHg = 198, quant_MMHg = 202, nat = 200, 5)

MMHg_concentrations_byinjection <- MMHg(peaks_1, masses_1, abundances_1, 24, 3, type = 'liquid', incub_IHg = 199, incub_MMHg = 201, quant_IHg = 198, quant_MMHg = 202, nat = 200, 0.22)

#or calculating the mean between the injections 
mean_IHg_concentration <- mean_IHg(peaks_1, masses_1, abundances_1, 24, 3, type = 'liquid', incub_IHg = 199, incub_MMHg = 201, quant_IHg = 198, quant_MMHg = 202, nat = 200, 5)

mean_MMHg_concentration <- mean_MMHg(peaks_1, masses_1, abundances_1, 24, 3, type = 'liquid',incub_IHg = 199, incub_MMHg = 201, quant_IHg = 198, quant_MMHg = 202, nat = 200, 0.22)


```

##Calculate methylation/demethylation rate constant 

The functions km() and mean_km() can be used to calculate directly the potential methylation rate constant by injection or by sample. 

```{r, methylation}
#Either calculating the individual values of each injection
km_byinjection <- km(peaks_1, masses_1, abundances_1, samples = 24, injections = 3, type = 'liquid', incub_IHg=199, incub_MMHg=201, quant_MMHg = 202, quant_IHg= 198, nat=200, spike_concentration_MMHg=0.22, spike_concentration_IHg=5)

#or the mean between the injections 
mean_km <- mean_km(peaks_1,masses_1,abundances_1,samples = 24 , injections = 3, type = 'liquid', incub_IHg=199, incub_MMHg=201, quant_MMHg = 202, quant_IHg= 198, nat=200, spike_concentration_MMHg=0.22, spike_concentration_IHg=5)

```

Similarly to the functioning of km() and mean_km(), the functions kd() and mean_kd() can be used to obtain the demethylation rate constant. 

```{r, demethylation}
#Either calculating the individual values of each injection
kd_byinjection <- kd(peaks_1, masses_1, abundances_1, samples = 24, injections = 3, type = 'liquid',incub_IHg=199, incub_MMHg=201, quant_MMHg = 202, quant_IHg= 198, nat=200, spike_concentration_MMHg=0.22, spike_concentration_IHg=5)

#or the mean between the injections 
mean_kd <- mean_kd(peaks_1,masses_1,abundances_1,samples = 24 , injections = 3, type = 'liquid', incub_IHg=199, incub_MMHg = 201, quant_MMHg = 202, quant_IHg= 198, nat=200, spike_concentration_MMHg=0.22, spike_concentration_IHg=5)

```


##References

- Monperrus, M., Tessier, E., Veschambre, S., Amouroux, D., & Donard, O. (2005). Simultaneous speciation of mercury and butyltin compounds in natural waters and snow by propylation and species-specific isotope dilution mass spectrometry analysis. Analytical and Bioanalytical Chemistry, 381(4), 854–862. https://doi.org/10.1007/s00216-004-2973-7
- Rodriguez-Gonzalez, P., Bouchet, S., Monperrus, M., Tessier, E., & Amouroux, D. (2013). In situ experiments for element species-specific environmental reactivity of tin and mercury compounds using isotopic tracers and multiple linear regression. Environmental Science and Pollution Research, 20(3), 1269-1280. https://doi.org/10.1007/s11356-012-1019-5
